State of the Art in Concurrent Priority Queues \\ Student Paper for Seminar in Algorithms 2013W \\ Technical University of Vienna

### 2014-03-17 paper/paper.tex
Extend Wimmer section

@@

Once the prefix of logically deleted nodes reaches a specified length \lstinline|BoundOffset|,
the first thread to observe this fact within \lstinline|DeleteMin| performs the actual physical
deletions by updating \lstinline|slist.head[0]| to point at the last logically deleted node with a
single \ac{FAO} operation. The remaining \lstinline|slist.head| pointers are then updated, and
all physically deleted nodes are marked for recycling.

@@

The body of work discussed in previous sections
creates the impression that limits of strict, deterministic priority queues have been reached.
In particular, \citeauthor{linden2013skiplist} conclude that scalability is solely limited by \lstinline|DeleteMin|,
and that less than one modified memory location per thread and operation would have to be read
in order to achieve improved performance \cite{linden2013skiplist}. % TODO: Cites for this entire section.

Recently, relaxation of provided guarantees have been investigated as another method of reducing
contention and improving disjoint-access parallelism.
For instance, k-FIFO queues \cite{} have achieved considerable
speed-ups compared to strict FIFO queues by allowing \lstinline|Dequeue| to return elements
up to k positions out of order.

@@

\citeauthor{wimmer2013data} presented the first relaxed, linearizable, and lock-free priority queue
in \cite{wimmer2013data}. An open-source implementation is available in the task-scheduling system
\emph{Pheet}\footnote{\url{http://pheet.org}}.

The \citeauthor{wimmer2013data} priority queue was evaluated using a label-correcting variant
of Dijkstra's shortest path algorithm \cite{}. Their model creates a task for each node
expansion, and therefore comes with a considerable task scheduling overhead. Nonetheless,
the parallel implementation scales well up to 10 threads, and further limited performance
gains are made until 40 threads. To date, no direct comparisons to other concurrent
priority queues have been possible since the data structure is strongly tied to \emph{Pheet},
but a separate implementation of k-priority queues is planned.

>>>>

% TODO: Mention inserting flag and memory management!

@@

Once the prefix of logically deleted nodes reaches a specified length {\lstset{breaklines,breakatwhitespace} \lstinline|BoundOffset|},
the first thread to observe this fact within \lstinline|DeleteMin| performs the actual physical
deletions by updating \lstinline|slist.head[0]| to point at the last logically deleted node with a
single \ac{FAO} operation. The remaining \lstinline|slist.head| pointers are then updated, and
all physically deleted nodes are marked for recycling.

@@

The body of work discussed in previous sections
creates the impression that the outer limits of strict, deterministic priority queues have been reached.
In particular, \citeauthor{linden2013skiplist} conclude that scalability is solely limited by \lstinline|DeleteMin|,
and that less than one modified memory location per thread and operation would have to be read
in order to achieve improved performance \cite{linden2013skiplist}.

Recently, relaxation of provided guarantees have been investigated as another method of reducing
contention and improving disjoint-access parallelism.
For instance, k-FIFO queues \cite{kirsch2012fast} have achieved considerable
speed-ups compared to strict FIFO queues by allowing {\lstset{breaklines,breakatwhitespace} \lstinline|Dequeue|} to return elements
up to $k$ positions out of order.

@@

\citeauthor{wimmer2013data} presented the first relaxed, linearizable, and lock-free priority queue
in \cite{wimmer2013data}. It is integrated as a priority scheduler into their \emph{Pheet} task-scheduling
system, and an open-source implementation is available\footnote{\url{http://pheet.org}}.

Their paper presents several variations on the common theme of priority queues: a distributed work-stealing
queue which can give no guarantees as to global ordering since it consists of separate priority queues
at each thread; a relaxed centralized priority queue in which no more than $k$ items are missed
by any processor; and a relaxed hybrid data structure which combines both ideas and provides
a guarantee that no thread misses more than $kP$ items where $P$ is the number of participating threads.
In this section, we examine only the hybrid variant since it provides both the scalability of work-stealing
queues and the ordering guarantees of the centralized priority queue.

\begin{figure}[h]
\begin{lstlisting}
struct globals_t {
  list_of_item_t global_list;
};

/* Thread-local items. */
struct locals_t {
  list_of_item_t local_list;
  pq_t prio_queue;
  size_t remaining_k;
};
\end{lstlisting}
\caption{\citeauthor{wimmer2013data} structure.}
\label{fig:wimmerq}
\end{figure}

The hybrid queue consists of a list of globally visible items and one local item list as well as a local
sequential priority queue per thread. The thread-local counter \lstinline|remaining_k| tracks how many more
items may be added to the local queue until all local items must be made globally visible to avoid
breaking guarantees.

Whenever an item is added, it is first added to both \lstinline|local_list| and \lstinline|prio_queue|
and \lstinline|remaining_k| is decremented. If \lstinline|remaining_k| reaches zero, then the local
item list is appended to the global list, and all not yet seen (by this thread) items of the global
queue are added to the local priority queue.

\lstinline|DeleteMin| simply pops the local priority queue as long is it is non-empty and the popped
item has already been deleted. When a non-deleted item is popped, it is atomically marked as deleted
and returned to the caller. If instead we are faced with an empty local queue, we attempt to spy,
i.e. copy items from another thread's local list.

The \citeauthor{wimmer2013data} priority queue was evaluated using a label-correcting variant
of Dijkstra's shortest path algorithms. Their model creates a task for each node
expansion, and therefore comes with a considerable task scheduling overhead. Nonetheless,
the parallel implementation scales well up to 10 threads, and further limited performance
gains are made until 40 threads. To date, no direct comparisons to other concurrent
priority queues have been possible since the data structure is strongly tied to \emph{Pheet},
but a separate implementation of k-priority queues is planned.
